Enforced Invariant Confluence

### 2018-03-29 doc/paper/sections/interactive_decision_procedure.tex
Finished interactive decision procedure section.

\section{An Interactive Decision Procedure}
% - describe how main theorem gives us intution about interactive dp
% - present the algorithm
% - walk through an example
% - describe how the algorithm is robust to errors

>>>>

\section{An Interactive Decision Procedure}\seclabel{InteractiveDecisionProcedure}
\newcommand{\Helper}{\textsf{Helper}}
\newcommand{\IsIclosed}{\textsf{IsIclosed}}
\newcommand{\IsInvConfluent}{\textsf{IsInvConfluent}}

\newcommand{\algocomment}[1]{\State \textcolor{flatdenim}{\texttt{//} #1}}
\begin{algorithm}[t]
  \caption{Interactive Invariant-Confluence Decision Procedure}%
  \algolabel{InteractiveDecisionProcedure}
  \begin{algorithmic}
    \Function{IsInvConfluent}{$O$, $s_0$, $T$, $I$}
    \State
    \Return $I(s_0)$ and
    \Call{Helper}{$O$, $s_0$, $T$, $I$, $\set{s_0}$, $\emptyset$}
    \EndFunction

    \State

    \algocomment{$R$ is a set of $\sTIreachable{}$ states.}
    \algocomment{$NR$ is a set of $\sTIunreachable{}$ states.}
    \algocomment{$I(s_0)$ is a precondition.}
    \Function{Helper}{$O$, $s_0$, $T$, $I$, $R$, $NR$}
    \State closed, $s_1$, $s_2$ $\gets$ \Call{IsIclosed}{$O$, $I - NR$}
    \If {closed}
      \State \Return true
    \EndIf
    \State Augment $R$ and $NR$ with random search and user input
    \If{$s_1, s_2 \in R$}
    \Return false
    \Else\
    \Return \Call{Helper}{$O$, $s_0$, $T$, $I$, $R$, $NR$}
    \EndIf
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\thmref{IclosureEquivalentIconfluence} tells us that if all
invariant-satisfying points are reachable, then invariant-closure and
invariant-confluence are equivalent. In this section, we present the
interactive invariant-confluence decision procedure shown in
\algoref{InteractiveDecisionProcedure} which takes advantage of this result.

\subsection{The Decision Procedure}
A user provides \algoref{InteractiveDecisionProcedure} with an object $O = (S,
\join)$, a start state $s_0$, a set of transactions $T$, and an invariant
$I$\footnote{%
  Formally, $I$ is a (possibly infinite) set of states (e.g. $\setst{(x, y) \in
  \ints \times \ints}{xy \leq 0}$), but in practice, our implementation of
  \IsInvConfluent{} takes in $I$ as a finite logical formula (e.g. $xy \leq
  0$).
}. The user then interacts with
\algoref{InteractiveDecisionProcedure} to iteratively eliminate unreachable
states from the invariant. Meanwhile, the algorithm leverages an
invariant-closure decision procedure to either (a) conclude that $O$ is or is
not \sTIconfluent{} or (b) provide counterexamples to the user to help them
eliminate unreachable states. After all unreachable states have been eliminated
from the invariant, \thmref{IclosureEquivalentIconfluence} allows us to reduce
the problem of invariant-confluence directly to the problem of
invariant-closure, and the algorithm terminates.
%
We now describe \algoref{InteractiveDecisionProcedure} in detail. An example of
how to use \algoref{InteractiveDecisionProcedure} on \exampleref{Z2} is given
in \figref{DecisionProcedureExample}.

\IsInvConfluent{} assumes access to an invariant-closure decision procedure
\IsIclosed$(O, I)$ which returns a triple (closed, $s_1$, $s_2$). closed is a
boolean indicating whether $O$ is \Iclosed{}. If closed is true, then $s_1$ and
$s_2$ are null. If closed is false, then $s_1$ and $s_2$ are a counterexample
witnessing the fact that $O$ is not \Iclosed{}. That is, $I(s_1)$ and $I(s_2)$,
but $\lnot I(s_1 \join s_2)$ (e.g. $s_1$ and $s_2$ from \exampleref{Z2}).  As
we mentioned earlier, we can (and do) implement the invariant-closure decision
procedure using an SMT solver like Z3~\cite{de2008z3}.

\IsInvConfluent{} first checks that $s_0$ satisfies the invariant. $s_0$ is
reachable, so if it does not satisfy the invariant, then $O$ is not
\sTIconfluent{} and \IsInvConfluent{} returns false. Otherwise,
\IsInvConfluent{} calls a helper function \Helper{} which---in addition to $O$,
$s_0$, $T$, and $I$---takes as arguments a set $R$ of $\sTIreachable{}$ states
and a set $NR$ of $\sTIunreachable{}$ states. Like \IsInvConfluent, \Helper$(O,
s_0, T, I, R, NR)$ returns whether $O$ is \sTIconfluent{} (assuming $R$ and
$NR$ are correct).  As \algoref{InteractiveDecisionProcedure} executes, $NR$ is
iteratively increased which removes unreachable states from $I$ until $I
\subseteq \setst{s}{\sTIreachable{s}}$.

First, \Helper{} checks to see if $O$ is \iclosed{(I - NR)}.
%
If \IsIclosed{} determines that $O$ is \iclosed{(I - NR)}, then by
\thmref{IclosureImpliesIconfluence}, $O$ is \sticonfluent{s_0}{T}{I-NR}, so
\[
  \setst{s \in S}{\stireachable{s_0}{T}{I-NR}{s}}
    \subseteq I - NR
    \subseteq I
\]
If $NR$ only contains $\sTIunreachable{}$ states, then the set of
$\sTIreachable{}$ states is equal to set of $\stireachable{s_0}{T}{I-NR}{}$
states which, as we just showed, is a subset of $I$.  Thus, $O$ is
\sTIconfluent{}, so \Helper{} return true.

If \IsIclosed{} determines that $O$ is \emph{not} \iclosed{(I - NR)}, then
we have a counterexample $s_1, s_2$. We want to determine whether $s_1$ and
$s_2$ are reachable or unreachable. We can do so in two ways.
%
First, we can randomly generate a set of reachable states and add them to
$R$. If $s_1$ or $s_2$ lies in $R$, then we know they are reachable.
%
Second, we can prompt the user to tell us directly whether or not the states
are reachable or unreachable.

In addition to labelling $s_1$ and $s_2$ as reachable or unreachable, the user
can also refine $I$ by augmenting $R$ and $NR$ arbitrarily (see
\figref{DecisionProcedureExample} for an example). In this step, we also make
sure that $s_0 \notin NR$ since we know that $s_0$ is reachable.

After $s_1$ and $s_2$ have been labelled as $\sTIreachable{}$ or not, we
continue. If both $s_1$ and $s_2$ are $\sTIreachable{}$, then so is $s_1 \join
s_2$, but $\lnot I(s_1 \join s_2)$. Thus, $O$ is not \sTIconfluent{}, so
\Helper{} returns false. Otherwise, one of $s_1$ and $s_2$ is
$\sTIunreachable{}$, so we recurse.

\Helper{} recurses only when one of $s_1$ or $s_2$ is unreachable, so $NR$
grows after every recursive invocation of \Helper{}. Similarly, $R$ continues
to grow as \Helper{} randomly explores the set of reachable states. As the user
sees more and more examples of unreachable and reachable states, it becomes
easier and easier for them to recognize patterns that define which states are
reachable and which are not. As a result, it becomes easier for a user to
augment $NR$ and eliminate a large number of unreachable states from the
invariant. Once $NR$ has been sufficiently augmented to the point that $I - NR$
is a subset of the reachable states, \thmref{IclosureEquivalentIconfluence}
guarantees that the algorithm will terminate after one more call to \IsIclosed.

\subsection{Tolerating User Error}
\algoref{InteractiveDecisionProcedure} is an \emph{interactive} decision
procedure. It requires that a user classify states as reachable or unreachable.
To err is human, so what happens when a user incorrectly classifies a state?
There are two possible errors that can be made, and
\algoref{InteractiveDecisionProcedure} can be made robust to both.

\textbf{A user can label an unreachable state as reachable.}
In this case, \Helper{} might erroneously find $s_1$ and $s_2 \in R$ and return
false, concluding that $O$ is not \sTIconfluent{} even when it is.  This is
inconvenient, but not catastrophic. We can modify
\algoref{InteractiveDecisionProcedure} so that \Helper{} requires that whenever
a user labels a state $s$ as $\sTIreachable{}$, they must also provide an
$\sTIreachable{}$ expression $e$ that evaluates to $s$. Here, $e$ acts a proof
that certifies that $s$ is indeed reachable. This increases the burden on the
user but completely eliminates this class of errors.

\textbf{A user can label a reachable state as unreachable.}
In this case, \IsIclosed$(O, I - NR)$ might return true, even though $O$ is not
\sTIconfluent. Thus, a user might falsely believe their distributed object to
be \sTIconfluent{} even though it isn't, and eventually one replica of their
distributed object might enter a state that violates the invariant.
%
We can mitigate this in two ways. First, we can aggressively expand $R$. If a
user ever labels a state $s$ as unreachable, but $s \in R$, we can notify the
user of their mistake. Second, \Helper{} returns true when $O$ is \iclosed{(I -
NR)} for some $NR$, so $O$ \emph{is} \sticonfluent{s_0}{T}{I-NR} (even though
it might not be \sTIconfluent{}). Thus, when a user replicates their
distributed object across a set of servers, they can deploy with the invariant
$I - NR$ instead of $I$. If $NR$ is correct and only contains unreachable
states, then deploying with $I - NR$ is equivalent to deploying with $I$. If
$NR$ is incorrect and contains some $\sTIunreachable{}$ states, then some of
these states are artificially made unreachable, but the system is still
guaranteed to never produce a state which violates $I$.

{\input{figures/decision_procedure_example}}
